How Slow do my Transient Typechecks Go?

### 2019-08-14 paper.tex
Copypasta evaluation - just need results now

@@

Moth is a virtual machine for Grace that we are developing as a
research platform \cite{roberts-and-co-ecoop-2019}. Like other VMs
based on the Truffle and Graal toolchain, Moth is a self-optimising
AST interpreter \cite{Wurthinger:2012:SelfOptAST}. 
The key idea is that an AST rewrites itself based on a program's run-time values
to reflect the minimal set of operations needed to execute the program
correctly. The rewritten AST is then compiled into efficient machine
code. This rewriting often depends on the dyanamic types of the
objects involved. In the simplest case, a ``self'' call (when one method
on an object invokes a second method on the exact same object) will
always result in executing the second method defined in same concrete
class as the first method. The second method can then be inlined into
the first method, avoiding the overheads of a machine-level subroutine
invocation and of an object-oriented dynamic dispatch.

@@

test protocol

@@

@@

   trying-to-be-cleverer results
  - single annotation overheads
  - using them go back to scatterplots (i.e. what you do in the latest graphs)

>>>>

\def\AWFY{Are\,We\,Fast\,Yet\xspace}

\usepackage{listings}

\lstdefinestyle{mystyle}{
    language=SAS,
    breakatwhitespace=true,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2,
    aboveskip=20pt,
    belowskip=10pt,
    xleftmargin=0.5cm,
    basicstyle=\footnotesize\ttfamily,
    commentstyle=\itshape\color{codegray},
    numberstyle=\footnotesize\color{codegray},
    stringstyle=\color{codegreen},
    keywordstyle = {\color{codepurple}},
    keywordstyle = [2]{\color{codered}},
    keywordstyle = [3]{\color{codered}},
    keywordstyle = [4]{\color{codegreen}},
    keywords={method,object,interface,type,var,def,return,class,for,in,if},
    otherkeywords = {:,->},
    morekeywords = [2]{:},
    morekeywords = [3]{->},
    morekeywords = [4]{true,false},
    morestring=*[d]{"},
    backgroundcolor={}
}

\lstset{style=mystyle}

@@

Moth is a virtual machine for Grace that we are developing as a
research platform \cite{roberts-and-co-ecoop-2019}. Like other VMs
based on the Truffle and Graal toolchain, Moth is a self-optimising
AST interpreter \cite{Wurthinger:2012:SelfOptAST}. 
The key idea is that an AST rewrites itself based on a program's run-time values
to reflect the minimal set of operations needed to execute the program
correctly. The rewritten AST is then compiled into efficient machine
code. This rewriting often depends on the dynamic types of the
objects involved. In the simplest case, a ``self'' call (when one method
on an object invokes a second method on the exact same object) will
always result in executing the second method defined in same concrete
class as the first method. The second method can then be inlined into
the first method, avoiding the overheads of a machine-level subroutine
invocation and of an object-oriented dynamic dispatch.

@@

In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the
performance of transient typechecks using the \AWFY
benchmark suite, plus other
benchmarks from the gradual-typing literature.
The goal was to complement our benchmarks with additional ones that are
used for similar experiments and can be ported to Grace.
To this end, we surveyed a number of papers\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
and selected benchmarks that have been used by multiple papers.
Some of these benchmarks overlapped with the \AWFY suite,
or were available in different versions.
While not always behaviorally equivalent,
we chose the \AWFY versions since we already used them to
establish the performance baseline.
The selected benchmarks as well as the papers in which they were used are shown in
\cref{tab:gradual-benchmarks}.

\begin{table}[htb]
  \caption{Benchmarks selected from literature.}
  \label{tab:gradual-benchmarks}
  \begin{center}
    \begin{tabular}{l l r}
      Fannkuch & \cite{Vitousek2017,Greenman2018} \\
      Float & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
      Go & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
      NBody & \cite{Kuhlenschmidt:2018:preprint,Vitousek2017,Greenman2018} & used \cite{Marr2016} \\
      Queens & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} & used \cite{Marr2016} \\
      PyStone & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
      Sieve & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} & used \cite{Marr2016} \\
      Snake & \cite{Takikawa2016,Muehlboeck2017,Bauman2017,Richards2017,Greenman2019jfp} \\
      SpectralNorm & \cite{Vitousek2017,Muehlboeck2017,Greenman2018} \\
    \end{tabular}
  \end{center}
\end{table}

The benchmarks were modified to have complete type information.  To
ensure correctness and completeness of these experiments, we added an
additional check to Moth that reports absent type information to
ensure each benchmark is completely typed.  To assess the performance
overhead of type checking, we compared the execution of Moth with all
checks disabled against an execution that has all checks enabled.  The
results of this evaluation demonstrated that Moth run programs with
transient typechecks enabled at an overhead of of 5\% (min. −13\%,
max. 79\%) over peak performance without typechecks.

We did not measure programs that mix typed and untyped code because
with our implementation technique a fully typed program is expected to
have the largest overhead. This is not the case for other kinds of
gradual type checking, which have given rise to more complex
evaluation techniques.  In particular, so-called ``Takikawa'' or
``Takikawa-Greenman'' evaluation \cite{takikawa2016,Greenman2019jfp}
benchmarks a \emph{lattice} of $23^N$ \emph{configurations} of each
benchmark. where a configuration is a particular mix of static and
dynamically typed code.

The Takikawa evaluation protocol was originally proposed for Typed
Racket, where static vs dynamic typing is set per module, so $N$ is
the number of modules. Grace, following other languages such as
Reticulated Python \cite{reticPython2014,monotonic2015,Vitousek2017},
allows programmers to choose typechecks at the level of individual
declarations: $N$ is the number of type annotations in the program.,
so checking an entire lattice for a benchmark of any size
is infeasible.  Vitousek et.al.\ modified the Takikawa protocol for
these kinds of langauges by taking an approach based on sampling
\cite{vitousek-transient-arXive-2019}.  The Takikawa-Vitousek protocol
divides the number of type annotations in an fully-typed program into
up to 100 intervals, and then randomly generates ten programs within
each interval by erasing type annotations.

To investigate Moth's support for transient typechecks further, we
carried out a (small?) Takikawa-Vitousek evaluation using (some of?)
our existing benchmarks.

\textbf{
!!!KJX: Need to update this paragraph with what we did this time around!!!.
To account for the complex warmup behavior
of modern systems\citep{Barrett:2017:VMW} as well as
the non-determinism caused by \eg garbage collection and cache effects,
we run each benchmark for \NumIterationsAll iterations in the same
VM invocation.\footnote{
For the Higgs VM, we only use \NumIterationsHiggs iterations,
because of its lower performance.
This is sufficient since Higgs's compilation approach induces less variation
and leads to more stable measurements.}
Afterwards, we inspected the run-time plots over the iterations
and manually determined a cutoff of \WarmupCutOff iterations for warmup,
\ie, we discard iterations with signs of compilation.
As a result, we use a large number of data points to compute the average,
but outliers, caused by \eg garbage collection, remain visible in the plots.
% In this work, we do not consider startup performance,
% because we want to assess the impact of gradual type checks
% on the best possible performance.
All reported averages use the geometric mean since they aggregate ratios.

% Yuria
%  - Ubuntu 16.04.4, Kernel 3.13
%  - 24 hyperthreads
%  - Intel Xeon E5-2620 v3 2.40GHz
% Graal 0.33 Feb. 2018
%
% TODO: update these details when we have time... not a rejection criterion,
% but needs to be correct in the final version
All experiments were executed on a machine running Ubuntu Linux 16.04.4,
with Kernel 3.13.
The machine has two Intel Xeon E5-2620 v3 2.40GHz,
with 6 cores each, for a total of 24 hyperthreads.
We used ReBench 0.10.1\citep{ReBench:2018}, Java 1.8.0\_171, Graal 0.33 (\code{a13b888}),
Node.js 10.4, and Higgs from 9 May 2018 (\code{aa95240}).
Benchmarks were executed one by one to avoid interference between them.
The analysis of the results was done with R 3.4.1,
and plots are generated with ggplot 2.2.1 and tikzDevice 0.11.
Our experimental setup is available online to enable reproductions.\footnote{
  \smtodo{merge changes, and tag final version}
\url{https://github.com/gracelang/moth-benchmarks}}
}

@@

give overall resuslts - one page of monochromatic scatterplots one for
each benchmark.

need to explKn how to read scatterplots by paraphrasing
\cite{vitousek-transient-arXive-2019}

@@

We hypothesised that the overhead of an individual type annotation may
be able to predict the performance of program configurations with that
annotation. To test this hypothesis, we constructed a series of
configurations, one for each type annotation, with only that
one annotation included.

   trying-to-be-cleverer results

   - single annotation overheads
  - using them go back to scatterplots (i.e. what you do in the latest graphs)

  - show that single annotation prediction doesn't reallty work\ldots
